The Perils of Exploration under Competition: \\ A Computational Modeling Approach

### 2020-07-09 journal paper/content/sim_details.tex
Added "prelims" subsections.
A quick pass on experiments.

\xhdr{Key Model Differences}
The timing of the model is the same as before where each firm commits to a MAB algorithm before the game starts and uses this algorithm to choose its actions. We focus on i.i.d. Bernoulli rewards: the  reward of each arm $a$ is drawn from $\{0,1\}$ independently with expectation $\mu(a)$. The mean rewards $\mu(a)$ are the same for all rounds and both firms, but initially unknown. However, instead of starting from some prior, we suppose that each firm has a uniform, ``fake", prior and that the initial information set of the firm is given by a ``warm start". Each algorithm receives a ``warm start": additional $T_0$ agents that arrive before the game starts, and interact with the firm as described above. The warm start ensures that each firm has a meaningful reputation and initial prior when competition starts..

In some of our experiments, one firm is the ``incumbent" who enters the market before the other (``late entrant"), and therefore enjoys a \emph{first-mover advantage}. Formally, the incumbent enjoys additional $X$ rounds of the ``warm start". We treat $X$ as an exogenous element of the model, and study the consequences for a fixed $X$.

\xhdr{MAB algorithms.} We consider the same three classes of bandit algorithms discussed in Section ~\ref{sec:model}, but take a representative algorithm from each class and look for qualitative differences between the different classes. We will utilize Thompson Sampling (\TS) from the ``smart" MAB algorithms, $\eps$-greedy (\DEG)\footnote{Throughout, we fix $\eps = 0.05$. Our pilot experiments showed that different $\eps$ did not qualitatively change the results.} from the ``naive" algorithms, and \DynGreedy (\DG). For ease of comparison, all three algorithms are parameterized with the same fake prior: namely, the mean reward of each arm is drawn independently from a $\Beta(1,1)$ distribution. Recall that Beta priors with 0-1 rewards form a conjugate family, which allows for simple posterior updates.

>>>>

\subsection{Experiment setup}
\label{expts-prelims}

%The timing of the model is the same as before where each firm commits to a MAB algorithm before the game starts and uses this algorithm to choose its actions. We focus on i.i.d. Bernoulli rewards: the  reward of each arm $a$ is drawn from $\{0,1\}$ independently with expectation $\mu(a)$. The mean rewards $\mu(a)$ are the same for all rounds and both firms, but initially unknown. However, instead of starting from some prior, we suppose that each firm has a uniform, ``fake", prior and that the initial information set of the firm is given by a ``warm start". Each algorithm receives a ``warm start": additional $T_0$ agents that arrive before the game starts, and interact with the firm as described above. The warm start ensures that each firm has a meaningful reputation and initial prior when competition starts..

\xhdr{Agent model.} All experiments use \HardMax response function (without mentioning it), except Section~\ref{sec:non_greedy} where we use \HardMaxRandom agents.

\xhdr{First-movers.}
Recall that in some of our experiments, one firm is the ``incumbent" who enters the market before the other (``late entrant"), and therefore enjoys a \emph{first-mover advantage}. Formally, the incumbent enjoys additional $X$ rounds of the ``warm start". We treat $X$ as an exogenous element of the model, and study the consequences for a fixed $X$.

\xhdr{MAB algorithms.} We consider the same three classes of bandit algorithms discussed in Section~\ref{sec:intro-model}, but take a representative algorithm from each class and look for qualitative differences between the different classes. We will utilize Thompson Sampling (\TS) from the ``adaptive-exploration" MAB algorithms, the dynamic $\eps$-greedy (\DEG)\footnote{Throughout, we fix $\eps = 0.05$. Our pilot experiments showed that different $\eps$ did not qualitatively change the results.} from the ``exploration-separating" algorithms, and the ``dynamic greedy" algorithm (\DG) as in Section~\ref{sec:theory-prelims}. For ease of comparison, all three algorithms are parameterized with the same ``fake" Bayesian prior: namely, the mean reward of each arm is drawn independently from a $\Beta(1,1)$ distribution. Recall that Beta priors with 0-1 rewards form a conjugate family, which allows for simple posterior updates.

\ascomment{TODO: add a bit more detail -- \eg define \eps-greedy.}

### 2020-07-09 journal paper/ec19paper/content/non_greedy_choice.tex
Reformatted so that: - all expts in the body are in one section. - all proofs in the app are in one section - all expts in the app are in one section.
A quick pass on experiments.

\section{Non-deterministic Choice Models}\label{sec:non_greedy}

Let us consider an extension in which the agents' choice is no longer deterministic. Recall that in our main model agents deterministically choose the firm with the higher reputation score; call this choice rule \emph{HardMax} (\HM). Now, we introduce some randomness: each agent selects between the firms uniformly with probability $\eps\in (0,1)$, and takes the firm with the higher reputation score with the remaining probability; call this choice rule \emph{HardMax with randomness} (\HMR).

One can view \HMR as a version of ``warm start", where a firm receives some customers without competition, but these customers are dispersed throughout the game. The expected duration of this ``dispersed warm start" is $\eps T$. If this quantity is large enough, we expect better algorithms to reach their long-term performance and prevail in competition. We confirm this intuition; we also find that this effect is negligible for smaller (but still relevant) values of $\eps$ or $T$.

@@

\begin{finding}\label{find:non_greedy_choice}
\textit{\TS is weakly dominant under the \HMR~ choice rule, if and only if $\eps T$ is sufficiently large. \asedit{\HMR leads to lower variance in market shares, compared to \HM.}}
\end{finding}

\footnotesize
\begin{table*}[t]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c||}{Heavy-Tail (\HMR with $\eps=.1$)}
  & \multicolumn{3}{c|}{Heavy-Tail (\HM)}\\
  \hline
  & \TS vs \DG & \TS vs \DEG  & \DG vs \DEG
 & \TS vs \DG & \TS vs \DEG  & \DG vs \DEG  \\
  \hline
$T = 2000$
 & \makecell{ \textbf{0.43} $\pm$ 0.02 \\Var: 0.15 }
  & \makecell{ \textbf{0.44} $\pm$ 0.02 \\Var: 0.15 }
  & \makecell{ \textbf{0.6} $\pm$ 0.02 \\Var: 0.1 }
  %%
 &  \makecell{ \textbf{0.29} $\pm$ 0.03 \\Var: 0.2 }
  & \makecell{ \textbf{0.28} $\pm$ 0.03 \\Var: 0.19 }
  & \makecell{ \textbf{0.63} $\pm$ 0.03 \\Var: 0.18 }
    \\
\hline
  $T= 5000$
   & \makecell{ \textbf{0.66} $\pm$ 0.01 \\Var: 0.056 }
  & \makecell{ \textbf{0.59} $\pm$ 0.02 \\Var: 0.092 }
  & \makecell{ \textbf{0.56} $\pm$ 0.02 \\Var: 0.098 }
  %%
 & \makecell{ \textbf{0.29} $\pm$ 0.03 \\Var: 0.2 }
 & \makecell{ \textbf{0.29} $\pm$ 0.03 \\Var: 0.2 }
 & \makecell{ \textbf{0.62} $\pm$ 0.03 \\Var: 0.19 }
 \\
  \hline
  $T = 10000$
  & \makecell{ \textbf{0.76} $\pm$ 0.01 \\Var: 0.026 }
 & \makecell{ \textbf{0.67} $\pm$ 0.02 \\Var: 0.067 }
 & \makecell{ \textbf{0.52} $\pm$ 0.02 \\Var: 0.11 }
 %%
  & \makecell{ \textbf{0.3} $\pm$ 0.03 \\Var: 0.21 }
  & \makecell{ \textbf{0.3} $\pm$ 0.03 \\Var: 0.2 }
  & \makecell{ \textbf{0.6} $\pm$ 0.03 \\Var: 0.2 }
  %%
  \\
   \hline
\end{tabular}
\normalsize
\caption{\HM and \HMR choice models on the Heavy-Tail MAB instance. Each cell describes the market shares in a game between two algorithms, call them Alg1 vs. Alg2, at a particular value of $t$. Line 1 in the cell is the market share of Alg 1: the average (in bold) and the 95\% confidence band.
%For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG.
Line 2 specifies the variance of the market shares across the simulations. The results reported here are with $T_0 = 20$.}
\label{tab:non_greedy_table}
\end{table*}

@@

\footnotesize
\begin{table*}[t]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c||}{Uniform (\HMR  with $\eps=.1$)}
  & \multicolumn{3}{c|}{Needle-In-Haystack (\HMR  with $\eps=.1$)}\\
  \hline
  & \TS vs \DG & \TS vs \DEG  & \DG vs \DEG
 & \TS vs \DG & \TS vs \DEG  & \DG vs \DEG  \\
 \hline
$T = 2000$
 & \makecell{ \textbf{0.42} $\pm$ 0.02 \\Var: 0.13 }
 & \makecell{ \textbf{0.45} $\pm$ 0.02 \\Var: 0.13 }
 & \makecell{ \textbf{0.49} $\pm$ 0.02 \\Var: 0.093 }
  %%
  & \makecell{  \textbf{0.55} $\pm$ 0.02 \\Var: 0.15 }
  & \makecell{  \textbf{0.61} $\pm$ 0.02 \\Var: 0.13 }
  & \makecell{  \textbf{0.46} $\pm$ 0.02 \\Var: 0.12 }
    \\
\hline
  $T= 5000$
 & \makecell{ \textbf{0.48} $\pm$ 0.02 \\Var: 0.089 }
 & \makecell{ \textbf{0.53} $\pm$ 0.02 \\Var: 0.098 }
 & \makecell{ \textbf{0.46} $\pm$ 0.02 \\Var: 0.072 }
  %%
 & \makecell{  \textbf{0.56} $\pm$ 0.02 \\Var: 0.13 }
 & \makecell{  \textbf{0.63} $\pm$ 0.02 \\Var: 0.12 }
 & \makecell{  \textbf{0.43} $\pm$ 0.02 \\Var: 0.11 }
 \\
  \hline
  $T = 10000$
& \makecell{ \textbf{0.54} $\pm$ 0.01 \\Var: 0.055 }
& \makecell{  \textbf{0.6} $\pm$ 0.02 \\Var: 0.073 }
& \makecell{  \textbf{0.44} $\pm$ 0.02 \\Var: 0.064 }
 %%
  & \makecell{ \textbf{0.58} $\pm$ 0.02 \\Var: 0.083 }
  & \makecell{ \textbf{0.65} $\pm$ 0.02 \\Var: 0.096 }
  & \makecell{ \textbf{0.4} $\pm$ 0.02 \\Var: 0.1 }
  %%
  \\
   \hline
\end{tabular}
\caption{\HMR choice model for Uniform and Needle-In-Haystack MAB instances.}
%Same semantics as in Table \ref{tab:non_greedy_table}.
\label{tab:additional_results}
\end{table*}
\normalsize

Table \ref{tab:non_greedy_table} shows the average market shares under
the \HM~ vs \HMR~ choice rule. In contrast to the \HM~ model,
  \TS becomes weakly dominant under the \HMR model, as $T$ gets
  sufficiently large. These findings hold across all problem
instances, see Table \ref{tab:additional_results} (with the same semantics as in Table \ref{tab:non_greedy_table}).

>>>>

\subsection{Non-deterministic choice Model (HardMaxRandom)}\label{sec:non_greedy}

Let us consider an extension in which the agents' response function \eqref{eq:model-f} is no longer deterministic. We focus on \HardMaxRandom model, where each agent selects between the firms uniformly with probability $\eps\in (0,1)$, and takes the firm with the higher reputation score with the remaining probability.

One can view \HardMaxRandom as a version of ``warm start", where a firm receives some customers without competition, but these customers are dispersed throughout the game. The expected duration of this ``dispersed warm start" is $\eps T$. If this quantity is large enough, we expect better algorithms to reach their long-term performance and prevail in competition. We confirm this intuition; we also find that this effect is negligible for smaller (but still relevant) values of $\eps$ or $T$.

@@

\begin{finding}\label{find:non_greedy_choice}
\textit{\TS is weakly dominant under \HardMaxRandom, if and only if $\eps T$ is sufficiently large. Moreover, \HardMaxRandom leads to lower variance in market share, compared to \HardMax.}
\end{finding}

\footnotesize
\begin{table*}[t]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c||}{Heavy-Tail (\HMR with $\eps=.1$)}
  & \multicolumn{3}{c|}{Heavy-Tail (\HM)}\\
  \hline
  & \TS vs \DG & \TS vs \DEG  & \DG vs \DEG
 & \TS vs \DG & \TS vs \DEG  & \DG vs \DEG  \\
  \hline
$T = 2000$
 & \makecell{ \textbf{0.43} $\pm$ 0.02 \\Var: 0.15 }
  & \makecell{ \textbf{0.44} $\pm$ 0.02 \\Var: 0.15 }
  & \makecell{ \textbf{0.6} $\pm$ 0.02 \\Var: 0.1 }
  %%
 &  \makecell{ \textbf{0.29} $\pm$ 0.03 \\Var: 0.2 }
  & \makecell{ \textbf{0.28} $\pm$ 0.03 \\Var: 0.19 }
  & \makecell{ \textbf{0.63} $\pm$ 0.03 \\Var: 0.18 }
    \\
\hline
  $T= 5000$
   & \makecell{ \textbf{0.66} $\pm$ 0.01 \\Var: 0.056 }
  & \makecell{ \textbf{0.59} $\pm$ 0.02 \\Var: 0.092 }
  & \makecell{ \textbf{0.56} $\pm$ 0.02 \\Var: 0.098 }
  %%
 & \makecell{ \textbf{0.29} $\pm$ 0.03 \\Var: 0.2 }
 & \makecell{ \textbf{0.29} $\pm$ 0.03 \\Var: 0.2 }
 & \makecell{ \textbf{0.62} $\pm$ 0.03 \\Var: 0.19 }
 \\
  \hline
  $T = 10000$
  & \makecell{ \textbf{0.76} $\pm$ 0.01 \\Var: 0.026 }
 & \makecell{ \textbf{0.67} $\pm$ 0.02 \\Var: 0.067 }
 & \makecell{ \textbf{0.52} $\pm$ 0.02 \\Var: 0.11 }
 %%
  & \makecell{ \textbf{0.3} $\pm$ 0.03 \\Var: 0.21 }
  & \makecell{ \textbf{0.3} $\pm$ 0.03 \\Var: 0.2 }
  & \makecell{ \textbf{0.6} $\pm$ 0.03 \\Var: 0.2 }
  %%
  \\
   \hline
\end{tabular}
\normalsize
\caption{\HardMax (\HM) and \HardMaxRandom (\HMR) choice models on the Heavy-Tail MAB instance. Each cell describes the market shares in a game between two algorithms, call them Alg1 vs. Alg2, at a particular value of $t$. Line 1 in the cell is the market share of Alg 1: the average (in bold) and the 95\% confidence band.
%For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG.
Line 2 specifies the variance of the market shares across the simulations. The results reported here are with $T_0 = 20$.}
\label{tab:non_greedy_table}
\end{table*}

@@

\footnotesize
\begin{table*}[t]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c||}{Uniform (\HMR  with $\eps=.1$)}
  & \multicolumn{3}{c|}{Needle-In-Haystack (\HMR  with $\eps=.1$)}\\
  \hline
  & \TS vs \DG & \TS vs \DEG  & \DG vs \DEG
 & \TS vs \DG & \TS vs \DEG  & \DG vs \DEG  \\
 \hline
$T = 2000$
 & \makecell{ \textbf{0.42} $\pm$ 0.02 \\Var: 0.13 }
 & \makecell{ \textbf{0.45} $\pm$ 0.02 \\Var: 0.13 }
 & \makecell{ \textbf{0.49} $\pm$ 0.02 \\Var: 0.093 }
  %%
  & \makecell{  \textbf{0.55} $\pm$ 0.02 \\Var: 0.15 }
  & \makecell{  \textbf{0.61} $\pm$ 0.02 \\Var: 0.13 }
  & \makecell{  \textbf{0.46} $\pm$ 0.02 \\Var: 0.12 }
    \\
\hline
  $T= 5000$
 & \makecell{ \textbf{0.48} $\pm$ 0.02 \\Var: 0.089 }
 & \makecell{ \textbf{0.53} $\pm$ 0.02 \\Var: 0.098 }
 & \makecell{ \textbf{0.46} $\pm$ 0.02 \\Var: 0.072 }
  %%
 & \makecell{  \textbf{0.56} $\pm$ 0.02 \\Var: 0.13 }
 & \makecell{  \textbf{0.63} $\pm$ 0.02 \\Var: 0.12 }
 & \makecell{  \textbf{0.43} $\pm$ 0.02 \\Var: 0.11 }
 \\
  \hline
  $T = 10000$
& \makecell{ \textbf{0.54} $\pm$ 0.01 \\Var: 0.055 }
& \makecell{  \textbf{0.6} $\pm$ 0.02 \\Var: 0.073 }
& \makecell{  \textbf{0.44} $\pm$ 0.02 \\Var: 0.064 }
 %%
  & \makecell{ \textbf{0.58} $\pm$ 0.02 \\Var: 0.083 }
  & \makecell{ \textbf{0.65} $\pm$ 0.02 \\Var: 0.096 }
  & \makecell{ \textbf{0.4} $\pm$ 0.02 \\Var: 0.1 }
  %%
  \\
   \hline
\end{tabular}
\caption{\HardMaxRandom (\HMR) choice model for Uniform and Needle-In-Haystack MAB instances.}
%Same semantics as in Table \ref{tab:non_greedy_table}.
\label{tab:additional_results}
\end{table*}
\normalsize

Table \ref{tab:non_greedy_table} shows the average market shares under
\HardMax vs \HardMaxRandom. In contrast to what happens under \HardMax,
  \TS becomes weakly dominant under \HardMaxRandom, as $T$ gets
  sufficiently large. These findings hold across all problem
instances, see Table \ref{tab:additional_results} (with the same semantics as in Table \ref{tab:non_greedy_table}).
